Survival Compass

### 2024-04-22 paper/paper.qmd
updated title, updated discussion + conclusion, added appendix.

---
title: "Survival Compass"
subtitle: "Statistical Insights into Lung Cancer Patients Journey Post Diagnosis"
author: 
  - Lexi Knight
thanks: "Code and data are available at: https://github.com/LexiKnight/Lung_Cancer/tree/main"
date: today
date-format: long 
abstract: "This study investigates the impact of pathologic stage and treatment modalities on lung cancer survival post-diagnosis. Analysis of patient data reveals significant correlation between pathologic stage, seeking treatment and survival outcomes. Notably, patients at advanced stages with metastases in distant sites beyond the lung, extensive lymph node involvement and tumors with extensive growth, invading nearby structures demonstrate lower survival rates. These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

## Treatment Type by Pathologic Stage
This scatter plot @fig-four provides a view of how pathologic stage correlates with the time to death among lung cancer patients, differentiated by treatment type. Each dot represents an individual patient, color-coded to represent whether they received pharmaceutical therapy or radiation therapy. On the x-axis we have days to death after diagnosis and on the y-axis we have the pathologic stages. There are considerably fewer patients part of stages I, II, III, and IIB at diagnosis compared to the other ones. As the time from diagnosis to death increases along the x-axis, we observe a general trend downwards in pathologic stage, indicating disease progression over time. The trend lines fitted for each treatment type further highlight this progression, with radiation therapy showing a steeper decline compared to pharmaceutical therapy. 
This illustrates that upon diagnosis, patients with more advanced pathologic stages are associated with a shorter survival time. Additionally, this suggests that patients receiving pharmaceutical therapy tend to exhibit a slower advancement in pathologic stage over time, thus increasing their survival time more so than those undergoing radiation therapy.

@@

## Treatment Type by Presence of Distant Metastasis
In this scatter plot, @fig-five we examine the relationship between the presence of distant metastasis, time to death, and treatment type for lung cancer patients. Similarly to figure above, the dots on the plot represent individual patients, color-coded based on their treatment regimen. Days to death after diagnosis is on the x-axis and on the y-axis we have the degree of the presence of distant metastasis. The majority of patients are diagnosed at classification 'MO' and hence are caught before cancer is spread from the primary site of origin. As patients survival time increases, excluding  category 'MX', there is a decrease in the presence of distant metastasis, denoted by the downward trend in the scatter plot. Patients undergoing pharmaceutical therapy show a sharper decrease in distant metastasis compared to those receiving radiation therapy. 
The trend line for radiation therapy is nearly a plateau, insinuating that radiation therapy was more effective in prolonging survival. This suggests that pharmaceutical therapy may  correlate with a lower likelihood of survival given distant metastasis over time compared to radiation therapy. Additionally, patients associated with higher classification of metastasis appear to have shorter survival times.

@@

# Display the plot
plot2
```
## Treatment Type by Lymph Node Involvement 
This scatter plot  @fig-six explores the relationship between lymph node involvement, time to death, and treatment type among cancer patients. Like above, each dot symbolizes an individual patient, with colors indicating their treatment type; pharmaceutical therapy or radiation therapy. Days to death is on the x-axis and lymph node involvement classifications are on the y-axis. At diagnosis, most patients are classified under either NO, N1 or N2. As survival time increases, there is an observable decline in lymph node involvement, reflected by the downward trend in the scatter plot. The trend lines further illuminate this pattern, with patients undergoing pharmaceutical therapy exhibiting a slightly steeper decline in lymph node involvement over time compared to those receiving radiation therapy, however this is very minimal and insignificant difference between the two treatment types. 
This suggests that pharmaceutical therapy and radiation therapy are both equally associated with increasing survival time based on lymph node involvement. Moreover, patients associated with higher classification of involvement with lymph nodes appear to have shorter survival times.

@@

# Display the plot
plot3
```
## Treatment Type by Tumor Size
In this scatter plot, @fig-seven we investigate the relationship between tumor size, time to death, and treatment type for cancer patients. Each dot on the plot corresponds to an individual patient, categorized by their treatment regimen whether it is pharmaceutical therapy or radiation therapy. The x-axis represents the time from diagnosis until death and the y-axis is comprised of the tumor sizes. At diagnosis most patients are classified with a T2 tumour. As survival time increases, there is a noticeable reduction in tumor size, depicted by the downwards trajectory in the scatter plot. The trend lines further accentuate this trend, with patients undergoing radiation therapy demonstrating a more pronounced decrease in tumor size over time compared to those receiving pharmaceutical therapy. 
This suggests that patients associated with larger tumors tend to have shorter survival times. This suggests that pharmaceutical therapy might be associated with a slower tumor growth rate and hence be more efficient in treating lung cancer patients in terms of tumor size.

@@

# Discussion {#sec-discussion}

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

@@

# Additional data details

# Model details {#sec-model-details}
we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

## Diagnostics
Is this needed?

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

>>>>

---
title: "Survival Compass"
subtitle: "Understanding the Impact of Pathologic Stage and Treatment Modalities on Lung Cancer Patients' Survival Post-Diagnosis."
author: 
  - Lexi Knight
thanks: "Code and data are available at: https://github.com/LexiKnight/Lung_Cancer/tree/main"
date: today
date-format: long 
abstract: "This study investigates the impact of pathologic stage and treatment modalities on lung cancer survival post-diagnosis. Analysis of patient data reveals significant correlation between pathologic stage, seeking treatment and survival outcomes. Notably, patients at advanced stages with metastases in distant sites beyond the lung, extensive lymph node involvement and tumors with extensive growth, invading nearby structures demonstrate lower survival rates. These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

## Treatment Type by Pathologic Stage
This scatter plot @fig-four provides a view of how pathologic stage correlates with the time to death among lung cancer patients, differentiated by treatment type. Each dot represents an individual patient, color-coded to represent whether they received pharmaceutical therapy or radiation therapy. On the x-axis we have days to death after diagnosis and on the y-axis we have the pathologic stages. There are considerably fewer patients part of stages I, II, III, and IIB at diagnosis compared to the other ones. As the time from diagnosis to death increases along the x-axis, we observe a general trend downwards in pathologic stage, indicating disease progression over time. The trend lines fitted for each treatment type further highlight this progression, with radiation therapy showing a steeper decline compared to pharmaceutical therapy.

@@

## Treatment Type by Presence of Distant Metastasis
In this scatter plot, @fig-five we examine the relationship between the presence of distant metastasis, time to death, and treatment type for lung cancer patients. Similarly to figure above, the dots on the plot represent individual patients, color-coded based on their treatment regimen. Days to death after diagnosis is on the x-axis and on the y-axis we have the degree of the presence of distant metastasis. The majority of patients are diagnosed at classification 'MO' and hence are caught before cancer is spread from the primary site of origin. As patients survival time increases, excluding  category 'MX', there is a decrease in the presence of distant metastasis, denoted by the downward trend in the scatter plot. Patients undergoing pharmaceutical therapy show a sharper decrease in distant metastasis compared to those receiving radiation therapy.

@@

# Display the plot
plot2
```
## Treatment Type by Lymph Node Involvement 
This scatter plot  @fig-six explores the relationship between lymph node involvement, time to death, and treatment type among cancer patients. Like above, each dot symbolizes an individual patient, with colors indicating their treatment type; pharmaceutical therapy or radiation therapy. Days to death is on the x-axis and lymph node involvement classifications are on the y-axis. At diagnosis, most patients are classified under either NO, N1 or N2. As survival time increases, there is an observable decline in lymph node involvement, reflected by the downward trend in the scatter plot. The trend lines further illuminate this pattern, with patients undergoing pharmaceutical therapy exhibiting a slightly steeper decline in lymph node involvement over time compared to those receiving radiation therapy, however this is very minimal and insignificant difference between the two treatment types.

@@

# Display the plot
plot3
```
## Treatment Type by Tumor Size
In this scatter plot, @fig-seven we investigate the relationship between tumor size, time to death, and treatment type for cancer patients. Each dot on the plot corresponds to an individual patient, categorized by their treatment regimen whether it is pharmaceutical therapy or radiation therapy. The x-axis represents the time from diagnosis until death and the y-axis is comprised of the tumor sizes. At diagnosis most patients are classified with a T2 tumour. As survival time increases, there is a noticeable reduction in tumor size, depicted by the downwards trajectory in the scatter plot. The trend lines further accentuate this trend, with patients undergoing radiation therapy demonstrating a more pronounced decrease in tumor size over time compared to those receiving pharmaceutical therapy.

@@

# Discussion {#sec-discussion}
Lung cancer presents significant challenges in oncology, with survival rates influenced by various factors. This study aimed to elucidate the interplay between pathologic stage, treatment modalities, and survival outcomes in lung cancer patients post-diagnosis.

## Survival Dynamics {#sec-first-point}
Eldridge explains that the severity of lung cancer is staged by pathologic stage which is determined by all three presence of distant metastasis, lymph node involvement and tumor size [@Eldrige]. Previous research found that patients with larger tumors and more extensive lymph node involvement are associated with a decrease in survival [@Zhang]. Additionally, greater presence of distant metastasis and thus advanced spread to other organs is shown to lower a patients survival [@Markman]. 
Our analysis revealed compelling associations between various clinical factors and survival times in lung cancer patients. Upon diagnosis, lung cancer patients with more advanced pathologic stages (@fig-four), classification of presence of distant metastasis (@fig-five), higher greater involvement of lymph nodes (@fig-six) and larger tumors (@fig-seven) were all shown to be associated with a shorter survival time.

## Treatment Efficacy {#sec-second-point}
An intriguing finding from our study was the differential impact of treatment modalities on survival outcomes. It was observed that @fig-four which looked at pathologic stage, demonstrates that patients receiving pharmaceutical therapy tend to exhibit a slower progression in pathologic stage over time, thus linked to having greater survival time compared to those undergoing radiation therapy. As pathologic stage is determined based on presence of distant metastasis, lymph node involvement and tumor size, this means that the main finding is that pharmaceutical therapy is the more effective treatment. In looking at these variables separately however, we see that not all the variables indicate that pharmaceutical therapy is superior.

Results looking at tumor size @fig-seven, pharmaceutical therapy is associated with slower tumor growth rate and thus pharmaceutical therapy is more potent at treating lung cancer patients, aligning with the results of pathologic stage @fig-four. On the other hand, the plateaued trend line for radiation therapy compared to the sharper decline of pharmaceutical therapy in @fig-five of presence of distant metastasis, this portrays radiation therapy was more effective in prolonging survival. Finally, with lymph node involvement @fig-six, pharmaceutical therapy and radiation therapy are both equally associated with increasing survival time based on lymph node involvement. These nuanced insights underscore the need for tailored treatment strategies based on individual patient characteristics and disease profiles.

## Limitations
Despite the insights gained from our study, several limitations warrant acknowledgment. As mentioned earlier, the data set included 981 patients and after removing patients with incomplete data, we were left with 382 patients. This is a much smaller sample and thus there is a chance of bias and impact validity of our findings. Another area of weakness could be that values for 'MX' presence of distant metastasis and 'NX" lymph node involvement which indicate cases that could not be assed, these were left in the data set as to not decrease the sample size anymore. It likely that had these been taken out, the results would be different. We made the executive decision to include these values as reasons for inability to be assessed could be because the spread was too small to be detected. Additionally, our study did not account for certain variables such as performance status, smoking status, and genetic mutations, which may influence treatment responses and survival outcomes. Another limitation is that this study treatment types were pharmaceutical and radiation treatment individually. In previous studies, pharmaceutical treatment is often used in combination with radiation therapy to treat lung cancer as its shown better results in destroying cancer cells. In some patients, pharmaceutical therapy ensures the tumor remains small in size and the radiation destroys the cancer.

## Future Directions
The dataset we used also offered information regarding demographics such as race, gender and year of birth. It would have been interesting to explore whether the treatment type and success would have been affected by these demographics. For example, radiation oncologists are looking into gender-dependent radiation therapies and are finding that female patients are more curative however they portray more side effects. Future research efforts should aim to address these gaps and explore the role of novel treatment modalities, biomarkers, and personalized medicine approaches in improving lung cancer prognosis.

# Conclusion {#sec-conclusion}
In conclusion, our study provides valuable insights into the complex factors influencing survival outcomes in lung cancer patients post-diagnosis. By elucidating the relationships between pathologic stage, treatment modalities, and survival times, we aim to inform clinical practice and guide future research endeavors. Although we established a general trend for those more likely to survive longer post-diagnosis, it's important to note that no two lung cancer patients are the same due to gene mutations present in the tumor and thus survival will vary person to person @Eldridge. Additionally, treatments are constantly improving and changing. Ultimately, our collective efforts must continue to focus on advancing early detection methods, refining treatment strategies, and enhancing the quality of care for lung cancer patients worldwide.

“words and phrases to avoid”: enduring; landscape or evolving landscape; intricate; meticulous approach; nuanced; robust; pivotal; crafting; crucial; layered perspective.
doing extensive/exhaustive statistical analysis/statistical endeavors on comprehensive/substantial data

@@

### Pathologic Stage ###
* stage is determined by tumor size, number of tumors found and where
  the cancer has spread
Stage I: cancer is found early, is in one lung, has not spread to LN. Localized
Stage II: found early but has spread inside or just ouside of one lung
  and may have spread to nearby LN on same side of the body. Regional spread.
Stage III: cancer found in one lung and has spread to the chest and LN
  further further away from the lungs. Regional spread.
Stage IV: advanced cancer and has spread to the lining or fluid around 
  the heart or lungs, or has spread to distant parts of the body. 
  Includes having tumors in both lungs, no matter the size of the tumors.

### Tumor size ###
TX: size cannot be assessed - too small for testing
T1: <3cm diameter, only in lungs 
T1a: <1cm, have not reached tissues surrounding lung/main branches of the
  airways
T1b: 1-2cm, have not reached tissues surrounding lung/main branches of the
  airways
T2: 3-5cm, grown into main bronchus/inner lining of lungs or caused lung 
  collapse or swelling. 
T2a: 3-4cm, grown into main airways and/or tissue around the lungs
T2b: 4-5cm, grown into main airways and/or tissue around the lungs
T3: 5-7cm, growth beyond primary site, into chest wall/ nearby site or
  other tumors in the same lobe of the lung
T4: >7cm, extensive tumor growth, invading nearby structures/organs. Growth
  into center of chest, to the diaphragm, heart or its major blood vessels,
  windpipe, esophagus, spine or other tumors in same lobe of the lung

### Presence of distant metastasis ###
MX: not evaluated 
M0: no distant metastasis, cancer not spread beyond site of origin/ nearby LN.
  Localized.
M1a: presence of metastasis to distant sites within the same lung/pleura.
  Regional.
M1b: metastasis to distant sites beyong lung and pleura, ex. brain, liver,
  bones. Distant.

### Lymph Node (LN) involvement ###
NX: LN cannot be assessed
N0: no regional LN metastasis, cancer not spread to LN
N1: involvement of regional LN
N2: more extensive LN involvement

### Treatment type ###
Pharmaceutical therapy 
Radiation Therapy (RT)

(AJCC, 2017)
Stage I = T1 to T2a, N0, M0
Stage II = T1 to T3, N0 to N1, M0
Stage III = T1 to T4, N0 to N3, M0
Stage IV = Any T, Any N, M1s
